Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-20 Manuscript/manuscript.Rmd
manuscript edits
small polishing

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary (Y=1), or not (Y=0). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  <!--When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D).--> As a normalization procedure, we performed a $log_{2}$ transformation on the distances prior to modelling (See Methods Section).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the Arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines $L = \{GM12878, K562\}$, at four different resolutions $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. For each classifier, we considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, four re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9 [??? In Methods, not here, you should explain why]), $C = \{1, 2, ..., 22\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials. [??? Currently, Methods and the results are redundant. Look how others present their models in the results section, and how methods section is structured in such cases. Specifically, some key definitions, like OC, OP, Distance should be solidly introduced in the results section, while log2 status of distance and supplementary figure 1 description are for methods]

## IMBALANCE RATIOS REMAINED SEVERE AND INVARIANT TOWARD RESOLUTION

@@

Using the Jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a Jaccard index range of (0-0.001) (Table 3). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high Jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

>>>>

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ is the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  <!--When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D).--> As a normalization procedure, we performed a $log_{2}$ transformation on the distances prior to modelling (See Methods Section). [??? Introduce all definitions in Results. Put all details in Methods. E.g., distance definition is crucial for understanding results, the fact of log2-transformation is Methods.]

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers [??? Do we need RF?] built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the Arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines $L = \{GM12878, K562\}$, at four different resolutions $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. For each classifier, we considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, four re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9 [??? In Methods, not here, you should explain why]), $C = \{1, 2, ..., 22\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials. [??? This is Methods. What is so compelling about those 'results' for the message you try to prove?] [??? Currently, Methods and the results are redundant. Look how others present their models in the results section, and how methods section is structured in such cases. Specifically, some key definitions, like OC, OP, Distance should be solidly introduced in the results section, while log2 status of distance and supplementary figure 1 description are for methods]

## IMBALANCE RATIOS REMAINED SEVERE irrespectively of RESOLUTION of Hi-C data

@@

Using the Jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a Jaccard index range of (0-0.001) (Table 3). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?] <!--The high Jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->
